---
title: 'Site- and gene-centric analysis'
output: rmarkdown::html_vignette
date: "`r BiocStyle::doc_date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>"
)
```

# Introduction

While 1, 2, and 3D pathway analyses are useful for data generated from experiments with different treatment/conditions, analysis designed for time-course data may be better suited to analysis experiments that profile multiple time points.

Here, we will apply `ClueR` which is an R package specifically designed for time-course proteomic and phosphoproteomic data analysis ([Yang et al. 2015])(https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004403).

# Loading packages and data

We will load the PhosR package with few other packages we will use for this tutorial.
```{r}
suppressPackageStartupMessages({
  library(parallel)
  library(ggplot2)
  library(ClueR)
  library(reactome.db)
  library(org.Mm.eg.db)
  library(annotate)
  library(PhosR)
})
```

We will load a dataset integrated from two time-course datasets of early and intermediate insulin signalling in mouse liver upon insulin stimulation to demonstrate the time-course phosphoproteomic data analyses.
```{r}
data("phospho_liverInsTC_RUV_pe")
ppe <- phospho.liver.Ins.TC.ratio.RUV.pe
ppe
```

# Gene-centric analyses of the liver phosphoproteome data

Let us start with gene-centric analysis. Such analysis can be directly applied to proteomics data. It can also be applied to phosphoproteomic data by using the `phosCollapse` function to summarise phosphosite information to proteins.

```{r}
# take grouping information
grps <- sapply(strsplit(colnames(ppe), "_"), 
            function(x)x[3])

# select differentially phosphorylated sites
sites.p <- matANOVA(ppe@assays@data$Quantification, 
                    grps)
ppm <- meanAbundance(ppe@assays@data$Quantification, grps)
sel <- which((sites.p < 0.05) & (rowSums(abs(ppm) > 1) != 0))
ppm_filtered <- ppm[sel,]

# summarise phosphosites information into gene level
ppm_gene <- phosCollapse(ppm_filtered, 
            gsub(";.+", "", rownames(ppm_filtered)), 
                stat = apply(abs(ppm_filtered), 1, max), by = "max")

# perform ClueR to identify optimal number of clusters
pathways = as.list(reactomePATHID2EXTID)

pathways = pathways[which(grepl("R-MMU", names(pathways), ignore.case = TRUE))]

pathways = lapply(pathways, function(path) {
    gene_name = unname(getSYMBOL(path, data = "org.Mm.eg"))
    toupper(unique(gene_name))
})

RNGkind("L'Ecuyer-CMRG")
set.seed(123)
c1 <- runClue(ppm_gene, annotation=pathways, 
            kRange = seq(2,10), rep = 5, effectiveSize = c(5, 100), 
            pvalueCutoff = 0.05, alpha = 0.5)

# Visualise the evaluation results
data <- data.frame(Success=as.numeric(c1$evlMat), Freq=rep(seq(2,10), each=5))
myplot <- ggplot(data, aes(x=Freq, y=Success)) + 
    geom_boxplot(aes(x = factor(Freq), fill="gray")) +
    stat_smooth(method="loess", colour="red", size=3, span = 0.5) +
    xlab("# of cluster") + 
    ylab("Enrichment score") + 
    theme_classic()
myplot

set.seed(123)
best <- clustOptimal(c1, rep=5, mfrow=c(2, 3), visualize = TRUE)
```

# Site-centric analyses of the liver phosphoproteome data

Phosphosite-centric analyses will perform using kinase-substrate annotation information from PhosphoSitePlus.

```{r}
RNGkind("L'Ecuyer-CMRG")
set.seed(1)
PhosphoSite.mouse2 = mapply(function(kinase) {
  gsub("(.*)(;[A-Z])([0-9]+;)", "\\1;\\3", kinase)
}, PhosphoSite.mouse)

# perform ClueR to identify optimal number of clusters
c3 <- runClue(ppm_filtered, annotation=PhosphoSite.mouse2, kRange = 2:10, rep = 5, effectiveSize = c(5, 100), pvalueCutoff = 0.05, alpha = 0.5)

# Visualise the evaluation results
data <- data.frame(Success=as.numeric(c3$evlMat), Freq=rep(2:10, each=5))
myplot <- ggplot(data, aes(x=Freq, y=Success)) + geom_boxplot(aes(x = factor(Freq), fill="gray"))+
  stat_smooth(method="loess", colour="red", size=3, span = 0.5) + xlab("# of cluster")+ ylab("Enrichment score")+theme_classic()
myplot

set.seed(1)
best <- clustOptimal(c3, rep=10, mfrow=c(2, 3), visualize = TRUE)

# Finding enriched pathways from each cluster
best$enrichList
```

# SessionInfo

```{r}
sessionInfo()
```


