---
title: 'Signalome construction with PhosR'
output: rmarkdown::html_vignette
date: "`r BiocStyle::doc_date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    warning = FALSE,
    message = FALSE,
    comment = "#>"
)
```

# Introduction

A key component of the `PhosR` package is to construct signalomes. The signalome construction is composed of two main steps: 1) kinase-substrate relationsip scoring and 2) signalome construction. This involves a sequential workflow where the outputs of the first step are used as inputs of the latter step. 

In brief, our kinase-substrate relationship scoring method (`kinaseSubstrateScore` and `kinaseSubstratePred`) prioritises potential kinases that could be responsible for the phosphorylation change of phosphosite on the basis of kinase recognition motif and phosphoproteomic dynamics. Using the kinase-substrate relationships derived from the scoring methods, we reconstruct signalome networks present in the data (`Signalomes`) wherin we highlight kinase regulation of discrete modules.

# Loading packages and data

First, we will load the `PhosR` package along with few other packages that we will be using in this section of the vignette.

```{r}
suppressPackageStartupMessages({
  library(PhosR)
  library(dplyr)
  library(ggplot2)
  library(GGally)
  library(ggpubr)
  library(calibrate)
  library(network)
})
```

We will also be needing data containing kinase-substrate annotations from `PhosphoSitePlus`, kinase recognition motifs from `kinase motifs`, and annotations of kinase families from `kinase family`. 
```{r}
data("KinaseMotifs")
data("KinaseFamily")
```

# Setting up the data

As before, we will set up the data by cleaning up the phoshophosite labels and performing RUV normalisation. We will generate the `ppe_RUV` matrix as in [batch_correction](https://pyanglab.github.io/PhosR/articles/batch_correction.html).

```{r}
data("phospho_L6_ratio_pe")
data("SPSs")

##### Run batch correction
ppe <- phospho.L6.ratio.pe
sites = paste(sapply(ppe@GeneSymbol, function(x)x),";",
                 sapply(ppe@Residue, function(x)x),
                 sapply(ppe@Site, function(x)x),
                 ";", sep = "")
grps = gsub("_.+", "", colnames(ppe))
design = model.matrix(~ grps - 1)
ctl = which(sites %in% SPSs)
ppe = RUVphospho(ppe, M = design, k = 3, ctl = ctl)

phosphoL6 = ppe@assays@data$normalised
```

# Generation of kinase-substrate relationship scores

Next, we will filtered for dynamically regulated phosphosites and then standardise the filtered matrix. 
```{r}
# filter for up-regulated phosphosites
phosphoL6.mean <- meanAbundance(phosphoL6, grps = gsub("_.+", "", colnames(phosphoL6)))
aov <- matANOVA(mat=phosphoL6, grps=gsub("_.+", "", colnames(phosphoL6)))
idx <- (aov < 0.05) & (rowSums(phosphoL6.mean > 0.5) > 0)
phosphoL6.reg <- phosphoL6[idx, ,drop = FALSE]

L6.phos.std <- standardise(phosphoL6.reg)
rownames(L6.phos.std) <- paste0(ppe@GeneSymbol, ";", ppe@Residue, ppe@Site, ";")[idx]
```

We next extract the kinase recognition motifs from each phosphosite. 
```{r}
L6.phos.seq <- ppe@Sequence[idx]
```

Now that we have all the inputs for `kinaseSubstrateScore` and `kinaseSubstratePred` ready, we can proceed to the generation of kinase-substrate relationship scores.
```{r fig.height=8, fig.width=8}
L6.matrices <- kinaseSubstrateScore(substrate.list = PhosphoSite.mouse, 
                                    mat = L6.phos.std, seqs = L6.phos.seq, 
                                    numMotif = 5, numSub = 1, verbose = FALSE)
set.seed(1)
L6.predMat <- kinaseSubstratePred(L6.matrices, top=30, verbose = FALSE) 
```

# Signalome construction

The signalome construction uses the outputs of `kinaseSubstrateScore` and `kinaseSubstratePred` functions for the generation of a visualisation of the kinase regulation of discrete regulatory protein modules present in our phosphoproteomic data. 
```{r fig.height=10, fig.width=10}
kinaseOI = c("PRKAA1", "AKT1")

Signalomes_results <- Signalomes(KSR=L6.matrices, 
                                predMatrix=L6.predMat, 
                                exprsMat=L6.phos.std, 
                                KOI=kinaseOI)
```

# Generate signalome map

We can also visualise the relative contribution of each kinase towards the regulation of protein modules by plotting a balloon plot. In the balloon plot, the size of the balloons denote the percentage magnitude of kinase regulation in each module. 
```{r fig.height=6, fig.width=15}
### generate palette
my_color_palette <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(8, "Accent"))
kinase_all_color <- my_color_palette(ncol(L6.matrices$combinedScoreMatrix))
names(kinase_all_color) <- colnames(L6.matrices$combinedScoreMatrix)
kinase_signalome_color <- kinase_all_color[colnames(L6.predMat)]

plotSignalomeMap(signalomes = Signalomes_results, color = kinase_signalome_color)
```

# Generate signalome network

Finally, we can also plot the signalome network that illustrates the connectivity between kinase signalome networks. 
```{r fig.height=5, fig.width=6}
plotKinaseNetwork(KSR = L6.matrices, predMatrix = L6.predMat, threshold = 0.9, color = kinase_all_color)
```

# SessionInfo

```{r}
sessionInfo()
```

